{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-08 thesis/tm.tex
abbreviate weak atomicity proof substantially, no ugly underlining

\newcommand\weakhilight[1]{{\ul{#1}}\xspace}

\revision{Prior work has highlighted the difference between
{\em strong} and {\em weak atomicity}
in TM systems
\cite{htm-subtleties,stm-weak-atomicity}.
Thus far we have assumed transactions will abort from any intervening memory conflict,
whether or not that access is itself transactional.
This corresponds to strong atomicity, the execution semantics provided by TSX.
% \cite{bugtm}. % sec 3.1 of that paper, not sure of an earlier citation that also states it in these exact terms
However, under weakly-atomic STM systems,
transactions do not abort when
a non-transactional conflicting access interleaves within them,
and condition \#1 of the execution semantics from \cref{sec:tm-design-failure} does not hold.
To model weak atomicity, we must weaken \Cref{lem:equiv} as follows (updated parts underlined for emphasis).
\vspace{1em}

\begin{lemma}[Equivalence of Aborts \weakhilight{under Weak Atomicity}]
	\label{lem:equiv-weak}
	For any interleaving prefix
	\[
	\begin{tabular}{c}
		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
	\end{tabular}
	\]
	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
	\begin{enumerate}
		\item \weakhilight{All events in
			$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},
			 \tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			 % ehh, bad parallel structure with below regarding the verb 'exists'
			are non-transactional, and the original interleaving itself exists, or}
		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
			(conflicting case), or
		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			(independent case)
	\end{enumerate}
	exists and is observationally equivalent.
\end{lemma}

\begin{proof}
	As before, \weakhilight{except first case on
	whether any intervening operations by $\tjj$/$\tkk$ are transactional.%
	\footnote{If $\tjj$/$\tkk$ already have a transaction active,
	their {\sf begin} must have occurred before $\tiiat{\alpha}$,
	and the theorem can apply starting from the earlier {\tt begin} with the thread IDs swapped.}
	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$ without aborting,
	and the new case \#1 applies.}
\end{proof}

@@

\begin{theorem}[\weakhilight{Weak} Atomicity of Transactions]
	\label{thm:atom-weak}
	For any state space $S$ of a transactionally-concurrent program,
	an equivalent state space exists in which all transactions are either executed atomically
	\weakhilight{with respect to other transactions only},
	or aborted immediately.
\end{theorem}

@@

Conveniently, in \cref{sec:tm-design-locks}'s formulation of
$I' = \exists m. \mathsf{lockify}_m(I)$,
\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
so mutual exclusion of $m$ is preserved,
%(a global lock does not protect against interleaving by arbitrary other code that never takes the lock either)
and
%the proof of
\Cref{thm:global-lock} applies unmodified.
\cref{sec:tm-impl-weak-atomicity}
discusses the implementation consequences of this weakening.}

>>>>

\revision{Prior work has highlighted the difference between
{\em strong} and {\em weak atomicity}
of transactions
\cite{htm-subtleties,stm-weak-atomicity}.
Thus far we have assumed transactions will abort from any intervening memory conflict,
whether or not that access is itself transactional.
This corresponds to strong atomicity, the execution semantics provided by TSX.
% \cite{bugtm}. % sec 3.1 of that paper, not sure of an earlier citation that also states it in these exact terms
However, weakly-atomic STM systems
do not abort transactions when
a non-transactional access conflicts,
and condition \#1 of the execution semantics from \cref{sec:tm-design-failure} does not hold.
To model weak atomicity, \Cref{lem:equiv} must be extended with a third possibility:
%as follows (updated parts underlined for emphasis).
\vspace{1em}

\begin{lemma}[Weak Equivalence of Aborts]
	\label{lem:equiv-weak}
	For any interleaving prefix such as defined in \Cref{lem:equiv}, either one of the original two conditions hold, or:
	\begin{enumerate}
		\setcounter{enumi}{2}
		\item All events in
			$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},
			 \tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			 % ehh, bad parallel structure with below regarding the verb 'exists'
			are non-transactional, and the original interleaving itself is a legal execution.
	\end{enumerate}
\end{lemma}

\begin{proof}
	As before, except first case on
	whether any intervening operations by $\tjj$/$\tkk$ are transactional.%
	\footnote{If $\tjj$ or $\tkk$ already have a transaction active,
	their {\sf begin} must have occurred before $\tiiat{\alpha}$,
	and the theorem can apply starting from the earlier {\tt begin} with the thread IDs swapped.}
	If so, either the original case \#1 or \#2 applies.
	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$
	(its preempted next transactional instruction)
	without aborting,
	and the new case \#3 applies instead.
\end{proof}

@@

\begin{theorem}[Weak Atomicity of Transactions]
	\label{thm:atom-weak}
	For any state space $S$ of a transactionally-concurrent program,
	an equivalent state space exists in which all transactions are either executed
	{\bf atomically
	with respect to other transactions only},
	or aborted immediately.
\end{theorem}

@@

Conveniently,
\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
so mutual exclusion of $m$ is preserved
in \cref{sec:tm-design-locks}'s formulation of
$I' = \exists m. \mathsf{lockify}_m(I)$,
%(a global lock does not protect against interleaving by arbitrary other code that never takes the lock either)
and
%the proof of
\Cref{thm:global-lock} applies unmodified.
\cref{sec:tm-impl-weak-atomicity}
discusses the implementation consequences of this weakening.}
